Designing Types for R, Empirically

### 2020-09-14 paper/main.tex
major edits, minor edit pass now

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{my_style}
\usepackage{listings, wrapfig,xspace}
\usepackage{paralist}
\usepackage{booktabs} % To thicken table lines

@@

\renewcommand{\null}{\texttt{\textbf{null}}\xspace}
\newcommand{\scalar}{\texttt{\textbf{scalar}}\xspace}
\renewcommand{\vector}{\texttt{\textbf{vector}}\xspace}
\newcommand{\class}{\texttt{\textbf{class}}\xspace}
\newcommand{\any}{\texttt{\textbf{any}}\xspace}
\newcommand{\environment}{\texttt{\textbf{env}}\xspace}
\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
\newcommand{\Language}{\texttt{\textbf{lang}}\xspace} %% Upper case!
\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
\newcommand{\NA}{\texttt{{NA}}\xspace}
\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
\newcommand{\T}{\ensuremath{T}\xspace}
\renewcommand{\S}{\ensuremath{S}\xspace}
\newcommand{\V}{\ensuremath{V}\xspace}
\newcommand{\A}{\ensuremath{A}\xspace}
\newcommand{\ID}{\ensuremath{ID}\xspace}
\newcommand{\F}{\ensuremath{F}\xspace}
\newcommand{\B}{\ensuremath{B}\xspace}
\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\VEC}[1]{#1\k{[]}\xspace}
\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}
\newcommand{\NASCA}[1]{\k{\string^}\!#1\xspace}

@@

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \any          & \it top type\\
  & |   & \null         & \it null type\\
  & |   & \environment  & \it environment type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
  & |   & \LIST\T                  & \it list type\\
  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T          & \it arguments \\
   & |   & \VARGS      & \vspace{5pt} \\
\V & ::= & ~\VEC\S     & \\
   & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
\S & ::= & \int        & \\
   & |   & \chr \\
   & |   & \dbl \\
   & |   & \lgl \\
   & |   & \clx \\
   & |   & \raw \vspace{5pt}\\
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

@@

\paragraph{Nullables} 
The \null type is inhabited by a singleton \code{NULL} value often used as a
sentinel or a stand in for arguments with no value.  To capture this
behavior, we introduce a nullable type \code{?}\T.  The difference between
the \NA and \code{NULL} values is that the latter cannot be stored in
vectors.

@@

\begin{figure}[h!tb]
\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
   $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
  $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
  \\
  \NAVEC\S  & <:& \VEC\S \\
  \T        & <:& \k{?}\T \\
  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
  \S        & <:& \VEC\S \\
  \lgl      & <:& \int\\
  \int      & <:& \dbl\\
  \dbl      & <:& \clx\\  
\end{tabular}
\caption{Simplification rules and subtyping}
\label{fig:siplification-rules}
\end{figure}

Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, immutable lists, and remove \null type to replace them with
nullables. For performance reasons, we bound lists to no more than five
types in their union, above that they become \LIST{any}.  Functions are alos
limited to twenty arguments. Higher-order functions are approximated by \any
to \any.

@@

\begin{table}[!h]
% BEGIN Autogenerated
\small\begin{tabular}{lrrrr}\toprule
Type & Args & \% of Args & Observations & \% of Obs.\\\midrule
\dbl               & 12,298 &     11.24 &  12,152,787 & 20.3 \\
\lgl               & 9,366 &      8.7 & 6,650,294 & 11.1 \\
\null              &  8,799 &      8.0 & 2,187,611 & 3.7 \\
\chr               &  8,727 &      8.0 & 2,564,207 & 4.3 \\
\VEC\dbl           &  7,190 &      6.6 &  4,934,773 & 8.2 \\
\VARGS             &  6,611 &      6.0 &  6,075,874 & 10.1 \\
\any               &  6,120 &      5.6 & 339,299 & 0.6 \\
\VEC\chr           &  4,325 &      4.0 &  1,060,466 & 1.8 \\
\CLASS{\texttt{matrix}}     & 4,152 & 3.8 & 2,805,718 & 4.7 \\
\CLASS{\texttt{data.frame}} & 2,608 & 2.4 &  352,655 & 0.6 \\
\bottomrule
\end{tabular}
\caption{Top types of arguments in R}\label{ctb}
\end{table}

@@

\AT{TODO: Update plots, reread and readjust wording and examples since some tables have changed.}

@@

The next step, is shown in Table~\ref{fig:singlefreq}, the table provides a
breakdown of types occuring in arguments without a union. Scalar, class and
vector are the most common type categories. The shaded rows correspond to
polymorphic types. Removing those gives us \CountMonomorphicPositions
(\PercMonomorphicPositions) monomorphic positions in a corpus of
\CountPositions parameters. With close to 65\% of monomorphic argument or
return values, it is fair to say that even a simple type language provides
significant benefits.

We would like to point out the number of occurrences of the singleton {\tt null} type, and discuss why we treat it as an {\it any}-type.
In R, it is common for programmers to include default values for arguments, and in many cases this value happens to be {\tt NULL}.
Our type analysis will report a {\tt null} type for these arguments if they are never passed a value during testing.
Thus, we interpret these as cases where we cannot be sure of the argument.

@@

If we look at the numbers from the point of view of functions and count how
many of their arguments are polymorphic, we observe that
\PercMonomorphicFunctions (\CountMonomorphicFunctions) functions are
monomorphic. The remaining \PercPolymorphicFunctions
(\CountPolymorphicFunctions) have at least one polymorphic parameter or
return type.  Figure~\ref{fig:funpoly} shows the distribution of functions
against the number of polymorphic arguments. Finally, we count
\CountMonomorphicPackages out of the 412 packages export only monomorphic
functions.

@@

\paragraph{NAs}
Our data supports making the presence of {\NA}s explicit. Only 1,058 (of
0.9\%) of arguments are marked as possibly having {\NA}s, thus the
overwhelming majority of types appear to be {\NA}-free.  In practice,
programmers check for them and sanitize them if they are present.  Consider
the \code{binom} package for computing confidence intervals and its
\code{binom.profile} function.  This attached code snippet highlights a data
sanitization pattern: the programmer first binds the vectors into a matrix,
then finds rows where both columns are not \NA, extracts non-\NA values and
stores them into \code{x} and \code{n} respectively.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  binom.profile <- function(x, n, conf.level=0.9, maxsteps=50,...) {
    xn <- cbind(x = x, n = n)
    ok <- !is.na(xn[, 1]) & !is.na(xn[, 2])
    x <- xn[ok, "x"]
    n <- xn[ok, "n"]
\end{lstlisting}

\paragraph{Scalars}
The data also suggests that programmers often use scalars, and do
dimensionality checks on their data.  In our data 31,533 (or 29\%) of the
arguments are scalar types. While not completely surprising, this is a
rather large number. Consider the \code{hankel.matrix} function, it takes
two arguments and checks that \code{n} is \int, that \code{x} is a vector,
and also, indirectly, that \code{n} is a a scalar (this comes from the fact
that it is used in the guard of a conditional which fails if \code{n} is
not a scalar).

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
 hankel.matrix <- function( n, x ) { 
  ### n = a positive integer value for the order of the Hankel matrix
  ### x = an order 2 * n + 1 vector of numeric values
    if ( n != trunc( n ) )  stop( "argument n ix not an integer" )
    if ( !is.vector( x ) )  stop( "argument x is not a vector" )
    m <- length( x )
    if ( m < n )  stop( "length of argument x is less than n" )
\end{lstlisting}

\paragraph{Nullables}
The number of argument which may be \c{NULL} is 3,165 (or 3.5\%). This is a
relatively small number of occurrences, but it is worth expressing the
potential for the presence of \c{NULL} as these would likely inhibit
optimizations.

\paragraph{Higher-Order Functions}
\typetracer assigns the type \CLASS{\texttt{function}} to function values.
The number of positions that receive a function is 1,831, which is just
2\% of all the positions for which we infer types. Given the small number of
occurrences, it is not worth complicating the inferred types with a complete
signature for these functions.

@@

\paragraph{Matrices}
Matrices are instances of the eponymous class. They have a \code{dims}
attribute indicating dimensions, and while not codified in the language
semantics, many internal functions coerce vectors to matrices automatically.
The \code{rowWeightedMeans} function calculates the weighted means of rows.
The programmer added a type check for \code{x}.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  rowWeightedMeans <- function(x, w=NULL, rows=NULL, cols=NULL, na.rm=FALSE, ...) {
    if (!is.matrix(x))   .Defunct(msg = sprintf("'x' should be a matrix.)
\end{lstlisting}

\paragraph{Data Frames}
One of the most popular classes in R is the \code{data.frame} class, making
up nearly 8\% of observed classes.  Data frames and the derivative
\code{tibble} and \code{data.table} types underpin many popular uses of R.
One way to deal with data frames is through the struct type, with a named
field for each column of the data frame, but as mentioned previously structs
introduced undue noise.  Further complicating data frames is that many
functions built to operate on them operate in a name-agnostic way.  For
instance, the \code{tidyverse} package ecosystem allows programmers to pass
column names to functions which operate on their data frames.  In base R,
idiomatic data frame use is to use string column names to select rows from
the frame (unless only a single column is of interest, wherein the \code{$}
syntax is appropriate).

@@

How robust are the inferred types? To measure this, we conducted another
large-scale experiment: for each package in the corpus, using the inferred type
signatures as contracts we ran all of the CRAN reverse dependencies for that
package. In total we ran \NUMPKGSEVAL unique packages and recorded
\NUMASSERTIONS total assertions. Overall, we found that only
\PERCFAILEDASSERTIONSWUNDEF of contract failed. The limit on number of arguments
(we record only 20) accounted for \PERCASSERTIONSUNDEF failed assertion. We
found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
never failed. The number of immaculate function types increases to
\PERCSUCCFUNNOSTHREE if we discount S3 object method dispatch. Overall, these
numbers are promising, and suggest that the type signatures are robust.

We breakdown the failed assertions by type in
Table~\ref{tbl:fail-breakdown-0}.  Accounting for 27\% of assertion
failures, is when \CLASS{\texttt{matrix}} is passed where a \VEC\dbl is
expected.  Considering these types, we might imagine them to be compatible,
however not allowing this coercion was a deliberate design decision:
coercion from vectors to matrices is ad hoc at best, and surely not a
practice codified in the language.  In a similar vein, another popular
failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
another case of commonly performed coercion.  We did not include these types
of coercions in our type annotation framework as programmers cannot rely on
them, and it is not always the case that the coercions are safe to perform.

\begin{table}
% BEGIN Autogenerated
\scriptsize
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 705,036 & 36.36 & 36.36 \\ 
\VEC\dbl & \VEC\int & 189,800 & 9.79 & 46.15 \\ 
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 5.16 & 51.31 \\
\CLASS{{\tt simpleError, error, condition}} & \CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \CLASS{{\tt randomForest}} | \VEC\dbl & 78,197 & 4.03 & 55.34 \\ 
  \CLASS{{\tt data.frame}} & \CLASS{{\tt data.frame, tbl, tbl\_df}} & 58,809 & 3.03 & 58.37 \\ 
  \CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 53,482 & 2.76 & 61.13 \\ 
 \VEC\dbl & \VEC\int & 33,350 & 1.72 & 62.85 \\ 
 \dbl & \CLASS{{\tt data.frame}} & 32261 & 1.66 & 64.52 \\ 
  \VEC\dbl & \CLASS{{\tt data.frame}} & 31361 & 1.62 & 66.13 \\ 
  \CLASS{{\tt xml\_node}} & \CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \CLASS{{\tt xml\_document, xml\_node}} & 30,330 & 1.56 & 67.70 \\ 
\bottomrule
\end{tabular}
% END Autogenerated
\caption{Top contract failures}
\label{tbl:fail-breakdown-0}
\end{table}

@@

Turning our attention now to arguments, we found that only
\PROPARGSFAILINGASSERTS of parameters failed.
Table~\ref{tbl:fail-breakdown-0} showed the runtime occurrences, but that
data alone does not tell the full story, as some failures may be
overrepresented if, e.g., a failing contract assertion was in a loop.  We were
interested in knowing for each of the most common violations in
Table~\ref{tbl:fail-breakdown-0}, how many different arguments had that
type, and how many of those exhibited the contract failure in question.
Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
folding away multiple identical failed contract assertions for the same
parameter position.  The first row of this table reads: a value of type
\VEC\dbl was passed to 15 different function parameters expecting a
\CLASS{\texttt{matrix}}, of which there are 1394 in total: 1\% of
\CLASS{\texttt{matrix}}-typed parameters were passed \VEC\dbl values.

We see that even though the double vector and matrix issue was wildly
prevalent in the raw, dynamic contract evaluation numbers, the number of
actual function argument types that were violated is very small.  The story
is similar with the double and integer coercion we mentioned earlier: it
represents many dynamic contract failures, but very few of the
\VEC\int-typed arguments have their contracts violated by \VEC\dbl-typed
values.  Rows 5 and 6 are interesting: we see that rather often arguments
expecting \CLASS{\texttt{timeSeries}} data are passed \VEC\dbl or
\CLASS{\texttt{matrix}} values.  This is a quirk of the \code{timeSeries}
package, whose functions often accept matrices and vectors, converting them
to time series in an ad hoc manner.  Note that the code coverage of the
\code{timeSeries} tests, examples, and vignettes on the \code{timeSeries}
package code is only 58\%, which is one possible explanation of why these
contract failures are occurring: the types that \typetracer generates are
only as good as the test code its run on.

>>>>

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{my_style}
\usepackage{listings, wrapfig,xspace}
\usepackage{paralist}
\usepackage{booktabs} % To thicken table lines
\usepackage{makecell}

@@

\newcommand{\nul}{\texttt{\textbf{null}}\xspace}
\newcommand{\scalar}{\texttt{\textbf{scalar}}\xspace}
\renewcommand{\vector}{\texttt{\textbf{vector}}\xspace}
\newcommand{\class}{\texttt{\textbf{class}}\xspace}
\newcommand{\any}{\texttt{\textbf{any}}\xspace}
\newcommand{\environment}{\texttt{\textbf{env}}\xspace}
\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
\newcommand{\Language}{\texttt{\textbf{lang}}\xspace} %% Upper case!
\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
\newcommand{\NA}{\texttt{{NA}}\xspace}
\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
\newcommand{\T}{\ensuremath{T}\xspace}
\renewcommand{\S}{\ensuremath{S}\xspace}
\newcommand{\V}{\ensuremath{V}\xspace}
\newcommand{\A}{\ensuremath{A}\xspace}
\newcommand{\ID}{\ensuremath{ID}\xspace}
\newcommand{\F}{\ensuremath{F}\xspace}
\newcommand{\B}{\ensuremath{B}\xspace}
\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\VEC}[1]{#1\k{[]}\xspace}
\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}
\newcommand{\NASCA}[1]{\k{\string^}\!#1\xspace}

@@

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \any          & \it top type\\
  & |   & \nul         & \it null type\\
  & |   & \environment  & \it environment type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
  & |   & \LIST\T                  & \it list type\\
  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T          & \it arguments \\
   & |   & \VARGS      & \vspace{5pt} \\
\V & ::= & ~\VEC\S     & \\
   & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
\S & ::= & \int        & \\
   & |   & \chr \\
   & |   & \dbl \\
   & |   & \lgl \\
   & |   & \clx \\
   & |   & \raw \vspace{5pt}\\
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

@@

\paragraph{Nullables} 
The \nul type is inhabited by a singleton \code{NULL} value often used as a
sentinel or a stand in for arguments with no value.  To capture this
behavior, we introduce a nullable type \code{?}\T.  The difference between
the \NA and \code{NULL} values is that the latter cannot be stored in
vectors.

@@

\begin{figure}[h!tb]
\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
   $\nul \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
  $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
  \\
  \NAVEC\S  & <:& \VEC\S \\
  \T        & <:& \k{?}\T \\
  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
  \S        & <:& \VEC\S \\
  \lgl      & <:& \int\\
  \int      & <:& \dbl\\
  \dbl      & <:& \clx\\  
\end{tabular}
\caption{Simplification rules and subtyping}
\label{fig:siplification-rules}
\end{figure}

Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, immutable lists, and remove \nul type to replace them with
nullables. For performance reasons, we bound lists to no more than five
types in their union, above that they become \LIST{any}.  Functions are alos
limited to twenty arguments. Higher-order functions are approximated by \any
to \any.

@@

\begin{table}[!h]
% BEGIN Autogenerated
\small\begin{tabular}{lrrrr}\toprule
Type & Args & \% of Args & Observations & \% of Obs.\\\midrule
\dbl               & 12,298 &     11.24 &  12,152,787 & 20.3 \\
\lgl               & 9,366 &      8.7 & 6,650,294 & 11.1 \\
\nul              &  8,799 &      8.0 & 2,187,611 & 3.7 \\
\chr               &  8,727 &      8.0 & 2,564,207 & 4.3 \\
\VEC\dbl           &  7,190 &      6.6 &  4,934,773 & 8.2 \\
\VARGS             &  6,611 &      6.0 &  6,075,874 & 10.1 \\
\any               &  6,120 &      5.6 & 339,299 & 0.6 \\
\VEC\chr           &  4,325 &      4.0 &  1,060,466 & 1.8 \\
\CLASS{\texttt{matrix}}     & 4,152 & 3.8 & 2,805,718 & 4.7 \\
\CLASS{\texttt{data.frame}} & 2,608 & 2.4 &  352,655 & 0.6 \\
\bottomrule
\end{tabular}
\caption{Top types of arguments in R}\label{ctb}
\end{table}

@@

@@

Table~\ref{fig:singlefreq} provides a
breakdown of types occurring in arguments without a union. Scalar, class and
vector are the most common type categories. The shaded rows correspond to
polymorphic types. 
When an argument's type is \nul, we say that the argument is polymorphic due to a limitation in our analysis:
In R, it is common for programmers to include default values for arguments, and in many cases this value happens to be {\tt NULL}.
Our type analysis will report a {\tt null} type for these arguments if they are {\it never passed a value} during testing.
We interpret these instances of \nul as polymorphic to capture that we cannot be sure of the actual type.

Removing the aforementioned instances of polymorphism gives us \CountMonomorphicPositions
(\PercMonomorphicPositions) monomorphic positions in a corpus of
\CountPositions parameters. With close to 60\% of monomorphic argument or
return values, it is fair to say that even a simple type language provides
significant benefits.

@@

If we look at the numbers from the point of view of functions and count how
many of their arguments are polymorphic, we observe that
\PercMonomorphicFunctions (\CountMonomorphicFunctions) functions are
monomorphic. The remaining \PercPolymorphicFunctions
(\CountPolymorphicFunctions) have at least one union or polymorphic parameter or
return type.  Figure~\ref{fig:funpoly} shows the distribution of functions
against the number of polymorphic arguments. Finally, we count that
\CountMonomorphicPackages out of the 412 packages export only monomorphic
functions.

@@

\paragraph{NAs}
Our data supports making the presence of {\NA}s explicit. Only 2923 (or
2.78\%) of arguments are marked as possibly having {\NA}s, thus the
overwhelming majority of types appear to be {\NA}-free.  In practice,
programmers check for them and sanitize them if they are present.  Consider
the \code{binom} package for computing confidence intervals and its
\code{binom.profile} function.  This attached code snippet highlights a data
sanitization pattern: the programmer first binds the vectors into a matrix,
then finds rows where both columns are not \NA, extracts non-\NA values and
stores them into \code{x} and \code{n} respectively.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  binom.profile <- function(x, n, conf.level=0.9, maxsteps=50, ...) {
    xn <- cbind(x = x, n = n)
    ok <- !is.na(xn[, 1]) & !is.na(xn[, 2])
    x <- xn[ok, "x"]
    n <- xn[ok, "n"]
    # ...
  }
\end{lstlisting}

\paragraph{Scalars}
The data also suggests that programmers often use scalars, and do
dimensionality checks on their data.  In our data 25,064 (or 33.33\%) of the
arguments are scalar types. While not completely surprising, this is a
rather large number. Consider the \code{hankel.matrix} function, it takes
two arguments and checks that \code{n} is \int, that \code{x} is a vector,
and also, indirectly, that \code{n} is a a scalar (this comes from the fact
that it is used in the guard of a conditional which fails if \code{n} is
not a scalar).

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily] 
  hankel.matrix <- function( n, x ) { 
    ### n = a positive integer value for the order of the Hankel matrix
    ### x = an order 2 * n + 1 vector of numeric values
    if ( n != trunc( n ) )  stop( "argument n ix not an integer" )
    if ( !is.vector( x ) )  stop( "argument x is not a vector" )
    m <- length( x )
    if ( m < n )  stop( "length of argument x is less than n" )
    # ...
  }
\end{lstlisting}

\paragraph{Nullables}
The number of argument which may be \c{NULL} is 5057 (or 4.44\%). This is a
relatively small number of occurrences, but it is worth expressing the
potential for the presence of \c{NULL} as these would likely inhibit
optimizations.

\paragraph{Higher-Order Functions}
\typetracer assigns the type \CLASS{\texttt{function}} to function values.
The number of positions that receive a function (possibly as part of a union) is 1,705, which is just
1.51\% of all the positions for which we infer types. Given the small number of
occurrences, it is not worth complicating the inferred types with a complete
signature for these functions.

@@

\paragraph{Matrices}
Matrices are instances of the eponymous class, representing 10.71\% of all classes occurring in types. They have a \code{dims}
attribute indicating dimensions, and while not codified in the language
semantics, many internal functions coerce vectors to matrices automatically.
The \code{rowWeightedMeans} function calculates the weighted means of rows.
The programmer added a type check for \code{x}.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  rowWeightedMeans <- function(x, w=NULL, rows=NULL, cols=NULL, na.rm=FALSE, ...) {
    if (!is.matrix(x))   .Defunct(msg = sprintf("'x' should be a matrix.)
    # ...
  }
\end{lstlisting}

\paragraph{Data Frames}
One of the most popular classes in R is the \code{data.frame} class, making
up  8.15\% of observed classes.  Data frames and the derivative
\code{tibble} and \code{data.table} types underpin much of the idiomatic usage of R.
One way to deal with data frames is through the struct type, with a named
field for each column of the data frame, but as mentioned previously structs
introduced undue noise.  Further complicating data frames is that many
functions built to operate on them operate in a name-agnostic way.  For
instance, the \code{tidyverse} package ecosystem allows programmers to pass
column names to functions which operate on their data frames.  In base R,
typical data frame use is to use string column names to select rows from
the frame (unless only a single column is of interest, wherein the \code{$}
syntax is appropriate).
In sum, data frames are a popular class of R values, and have spawned many derivative data types, such as tibbles and data tables.
We include a \CLASS{{\tt data.frame}} type to cover most use-cases, and we leave a richer type for future work on a full fledged object-oriented type system.

@@

We now ask {\it how robust are the inferred types}? To measure this, we conducted another
large-scale experiment: for each package in the corpus, using the inferred type
signatures as contracts we ran all of the CRAN reverse dependencies for that
package. In total we ran \NUMPKGSEVAL unique packages and recorded
\NUMASSERTIONS total assertions. Overall, we found that only
\PERCFAILEDASSERTIONSWUNDEF of contract failed. The limit on number of arguments
(we record only 20) accounted for \PERCASSERTIONSUNDEF failed assertion. We
found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
never failed. The number of immaculate function types increases to
\PERCSUCCFUNNOSTHREE if we discount S3 object method dispatch. Overall, these
numbers are promising, and suggest that the type signatures are indeed robust.

We breakdown the failed assertions by type in
Table~\ref{tbl:fail-breakdown-0}.  Accounting for 36.36\% of assertion
failures, is when a \VEC\dbl is passed where a \CLASS{\texttt{matrix}} is
expected.  Considering these types, we might imagine them to be compatible,
however not allowing this coercion was a deliberate design decision:
coercion from matrices to vectors is ad hoc at best, and surely not a
practice codified in the language. For instance, if it is determined that we should coerce the matrix it would need to be flattened, and should this be done row-wise or column-wise?

In a similar vein, another popular
failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
another case of commonly performed coercion.  We did not include these types
of coercions in our type annotation framework as programmers cannot rely on
them, and it is not always the case that the coercions are safe to perform.

The second row of the table is exemplary of a pattern where vectors are passed when scalars are expected.
In these cases, the functions exhibiting these assertion failures were under-tested, and can operate just as well on vectors of values.
As an example, this failure occurred in functions from the \code{lubridate} package which provides date/time functionality.
Many functions, e.g., \code{date_decimal} and \code{make_datetime} turn doubles into \CLASS{{\tt POSIXct, POSIXt}} (which are dates in R), and they can easily operate on vectors of doubles, producing lists of dates.

Finally, we point out that assertion failures of, e.g., \CLASS{{\tt data.frame}} values being passed to \CLASS{{\tt data.frame, tbl, tbl\_df}} arguments and \CLASS{{\tt xml\_node}} values being passed to an argument expecting other XML-like classes are related to our simplified take on class types. 
Our type system does not encode user-defined subtyping and coercion, which could help address these mismatches.

\begin{table}
% BEGIN Autogenerated
\scriptsize
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumul. \%\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 705,036 & 36.36 & 36.36 \\ 
\VEC\dbl & \dbl & 189,800 & 9.79 & 46.15 \\ 
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 5.16 & 51.31 \\
\CLASS{{\tt simpleError, error, condition}} & \makecell[tl]{\CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \\ \CLASS{{\tt randomForest}} | \VEC\dbl} & 78,197 & 4.03 & 55.34 \\ 
  \CLASS{{\tt data.frame}} & \CLASS{{\tt data.frame, tbl, tbl\_df}} & 58,809 & 3.03 & 58.37 \\ 
  \addlinespace
  \CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 53,482 & 2.76 & 61.13 \\ 
 \VEC\dbl & \VEC\int & 33,350 & 1.72 & 62.85 \\ 
 \dbl & \CLASS{{\tt data.frame}} & 32261 & 1.66 & 64.52 \\ 
  \VEC\dbl & \CLASS{{\tt data.frame}} & 31361 & 1.62 & 66.13 \\ 
  \CLASS{{\tt xml\_node}} & \makecell[tl]{\CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \\ \CLASS{{\tt xml\_document, xml\_node}}} & 30,330 & 1.56 & 67.70 \\ 
\bottomrule
\end{tabular}
% END Autogenerated
\caption{Top contract failures}
\label{tbl:fail-breakdown-0}
\end{table}

@@

Turning our attention now to arguments, we found that only
\PROPARGSFAILINGASSERTS of parameters failed.
Table~\ref{tbl:fail-breakdown-0} showed the runtime occurrences, but that
data alone does not tell the full story, as some failures may be
overrepresented if, e.g., a failing contract assertion was in a loop.  We were
interested in knowing for each of the most common violations in
Table~\ref{tbl:fail-breakdown-0}, how many different arguments had that
type, and how many of those exhibited the contract failure in question.
Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
folding away multiple identical failed contract assertions for the same
parameter position.  The first row of this table reads: a value of type
\VEC\dbl was passed to 15 different function parameters expecting a
\CLASS{\texttt{matrix}}, of which there are 1522 in total: 18 (1.18\%) of these
\CLASS{\texttt{matrix}}-typed parameters were passed \VEC\dbl values.

We see that even though the double vector and matrix issue was wildly
prevalent in the raw, dynamic contract evaluation numbers, the number of
actual function argument types that were violated is very small.  The story
is similar with the double and integer coercion we mentioned earlier: it
represents many dynamic contract failures, but very few of the
\VEC\int-typed arguments have their contracts violated by \VEC\dbl-typed
values.  Row six is interesting: we see that rather often arguments
expecting \CLASS{\texttt{timeSeries}} data are passed
\CLASS{\texttt{matrix}} values.  This is a quirk of the \code{timeSeries}
package, whose functions often accept matrices and vectors, converting them
to time series in an ad hoc manner.  Note that the code coverage of the
\code{timeSeries} tests, examples, and vignettes package code is only 58\%, which is one possible explanation of why these
contract failures are occurring: the types that \typetracer generates are
only as good as the test code its run on.
